---
title: "Joshua French"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_notebook
---

```{css, include=FALSE}
.action {
  # margin: 2em;
  padding: 2em;
  border: 1px solid grey;
  border-radius: 5px;
  background: #eeeeee;
}
```

# Crash Course in R Markdown

## What is R Markdown?

R Markdown is a tool for dynamic, reproducible creation of data-centric documents that include text, code, the results of executed code.

* R Markdown was created by [R Studio](https://www.rstudio.com) in 2014 (or at least that's the oldest discussion I can find).
* It is a variant of Markdown, which was created by John Gruber (of [Daring Fireball](https://daringfireball.net/) fame) and Aaron Swartz (co-creator of [Reddit](https://www.reddit.com)) as a lightweight markup language for creating formatted text using a plain-text editor.
* In general, a markdown language is a language that is written in plain-text but that is rendered different based on "markups" in the text.
* Has some capabilities similar to [Jupyter](https://jupyter.org/) notebooks.

R Markdown is a great way to:

* Create a data analysis notebook that documents how and why you did analysis a certain way.
* Share your work with others in a variety of formats.
* Make your work reproducible.

R Markdown is a very useful tool for the data scientist.

R Studio provides R Markdown (and other) cheatsheets at [https://rstudio.com/resources/cheatsheets/](https://rstudio.com/resources/cheatsheets/)

### Your turn
The easiest way to create a new R Markdown file is File $\to$ New File $\to$ R Markdown. Do this on your computer to see what is produced.

## What's in an R Markdown document?
An R Markdown document has three components:

1. An optional YAML header.
2. Formatted text.
3. **Chunks** of computer code.

Since the YAML header is option, we'll talk about it last.

## Formatting text in R Markdown.

### Headings and sections
Headings and sections of different levels are created with `#`.

* `#` indicates the first level
* `##` being the second level
* etc.

### Font styles and effects

* *Italic text*: `*text*` or `_text_`.
* **Bold text**: `**text**` or `__text__` (two underscores on each side)
* `Typewriter (code) text`: `` `text` ``
* Endash (--), emdash (--): `--` and `---`.
* ^superscript^ and ~subscript~: `^superscript^` and `~subscript~`
* ~~strikethrough~~: `~~strikethrough~~`
* Escaping special characters \*, \_, \$, \\, `` ` ``: `\*`, `\_`, `\$`,  `\\`, ``` `` ` `` ```
* Footnote^[This is a footnote]: `Footnote^[This is a footnote]`

Math is rendered using $\LaTeX$ (which is [pronounced](https://www.latex-project.org/about/) "Lah-tech" or "Lay-tech"). So you can use \$ and \$\$ to render inline math and display math like you would in $\LaTeX$.

E.g., Typing "The regression coefficient \$\\beta\$" will produce "The regression coefficient $\beta$".

### Your turn {.action}

1. Create a new R Script (Ctrl/Cmd + Shift + n).
2. Replace the file contents with the code below:
```{r, eval=FALSE}
---
title: "Bad jokes"
output: html_notebook
---

# First level

## Second level

### Third level

*This is italic text*

**This is bold text**

`This looks like computer code`

What's the difference between endash--endash and emdash---emdash?

How can I add a superscript to this? Like this: this^superscript^?

That's a joke if you didn't get the ~subtext~. 

Wow, two bad jokes in a row. I'd like to strike ~~that~~ from my memory.

I wish I could \*escape\* these bad jokes.

But I guess this will just be an unpleasant footnote^[unpleasand footnote].
```
3. Save the file (Ctrl/Cmd + s) with the name "bad_jokes.Rmd".
4. Click the Preview button and view the results.

## Numbered and unordered lists

### Numbered lists
Numbered lists with sub-bullets:

```{r, eval=FALSE}
A numbered list with subnumbering.

1. Item 1
    a. Sub-item 1 # two tabs, not one
    b. Sub-item 2
        i. Sub-sub-item 1 # four tabs, not two
2. Item 2
    a. Sub-item 1
```

Result:
Some text with bullets.

1. Item 1
    a. Sub-item 1
    b. Sub-item 2
        i. Sub-sub-item 1
2. Item 2
    a. Sub-item 1

Note: There must be a line between any text and the number `1`.

### Unordered lists

Unordered lists with sub-bullets:

```{r, eval=FALSE}
A numbered list with subnumbering.

1. Item 1
    a. Sub-item 1 # two tabs, not one
    b. Sub-item 2
        i. Sub-sub-item 1 # four tabs, not two
2. Item 2
    a. Sub-item 1
```

Result:
Some text with bullets.

1. Item 1
    a. Sub-item 1
    b. Sub-item 2
        i. Sub-sub-item 1
2. Item 2
    a. Sub-item 1

Note: There must be a line between any text and the \*.

### Your turn

Create an example of a numbered list with 4 levels and an unordered list with 3 levels.

## Code chunks

Chunks of code can be added in an R Markdown by placing the relevant code between ```` ```{r} ```` and ```` ``` ````. e.g., 

````
```{r, label = chunk_name}`r ''`
1 + 1
```
````

You can generate a code chunk template using Ctrl/Cmd + Alt + i.

The `chunk_name` is optional but allows you give give chunks names labels that you can refer back to. This is useful for complicated documents and reports. (See [Verbatim Code Chunks](https://bookdown.org/yihui/rmarkdown-cookbook/verbatim-code-chunks.html) for how to get the above code to render.)

Note: each label should be unique within the document. I don't recommend adding chunk labels unless you actually need them, otherwise you are likely to reuse a chunk label and make a headache for yourself!

### Chunk options
Code chunks have many options that can be used to customize the code. We will cover many important ones below.  I will specify the default option in parentheses, (), after the option. So this is what will be used by default if you don't specify anything.

Yuhui Xie, creator of **knitr**, provides a [complete list])(https://yihui.org/knitr/options/) of chunk options on his [website](https://yihui.org/).

### Code evaluation

* `eval`: (`TRUE`) A logical value indicating whether the chunk should be evaluated in the Console.
  * You can also evaluate **selected lines** if you provide a numeric vector. e.g., `eval = 1:3` would evaluate lines 1 through 3. 

### Text output

* `echo`: (`TRUE`). A logical value indicating whether the source code should be displayed.
  * You can also echo **selected lines** if you provide a numeric vector. e.g., `echo = 1:3` would echolines 1 through 3
* `results`: (`'markup'`). Determines hwo the results are displayed.
  * Other options are `'asis'`, `'hold'`, `'hide'`.
* `collapse`: (`FALSE`). A logical value indicating whether the output should be collapsed.
* `include`: (`TRUE`) A logical value indicating whether to include the chunk output.    * Set to `FALSE` when you want something to run but not be shown.

Other options: `warning`, `error`, `messages` (TRUE) determine whether warnings, errors, and messages will be displayed, respectively.

### Code styling

* `highlight`: (TRUE) A logical value indicating whether the source code syntax should be highlighted with different colors.
* `comment`: (`##`) The text that will be shown prior to evaluated code in the output block. 
  * Use `##` or `#>` to make the code blocks easier to copy and paste into the Console. `#>` is closer to what the Console output actually looks like.
* `tidy`: (FALSE) A logical value indicating whether the source code be automatically reformatted.
* `tidy.opts`: (NULL) A list of options that can be used to customize the `tidy` option above.
  * e.g., `tidy.opts = list(width.cutoff = 60)` controls the length of the code and trys to cute the code lines at 60 characters.

### Plots

* `fig.show`: (`'asis'`). Instructions for displaying plots. 
  * `'asis'`: Show plots in the order and places they would be if executed in the Console.
  * `'hold'`: Output all plots at the end of the chunk.
  * `'hide'`: Hide all plots.
  * `'animate'`: Combine and animate the plots
* `fig.align`: (`'default'`). A character vector specifying how to align the figures.
  * Allowablve values are: `'default'`, `'left'`, `'right'`, and `'center'`.
* `fig.width`, `fig.height`: (7). A numeric value indicating the width and height of the result figures.
* `fig.cap`: (NULL). A caption for the figure.


### Your turn

Copy and paste the following code into an R Markdown document and then either knit or preview it to see what you get.

````
# Evaluation

`eval = TRUE`
```{r, eval = TRUE}`r ''`
1 + 1
```

`eval = FALSE`
```{r, eval = FALSE}`r ''`
1 + 1
```

```{r, eval = 1:2}`r ''`
1 + 1
2 + 2
3 + 3
```

```{r, eval = -c(1, 3)}`r ''`
1 + 1
2 + 2
3 + 3
```

# Text output

## Echo
`echo = TRUE`
```{r, echo = TRUE}`r ''`
1:3
```

`echo = FALSE`
```{r, echo = FALSE}`r ''`
1:3
```

`echo = 1:2`
```{r, echo = 1:2}`r ''`
1:3
4:6
7:9
```

`echo = -1`
```{r, echo = -1}`r ''`
1:3
4:6
7:9
```

### results

`results = 'markup'`
```{r, results = 'markup'}`r ''`
1:3
4:6
7:9

`results = 'asis'`
```{r, results = 'asis'}`r ''`
1:3
4:6
7:9
```

`results = 'hold'`
```{r, results = 'hold'}`r ''`
1:3
4:6
7:9
```

`results = 'hide'`
```{r, results = 'hide'}`r ''`
1:3
4:6
7:9
```

````


### Options not discussed

* Caching: This has to do with saving previous code that has been evaluated.






