---
title: "Data frame manipulation with dplyr and tidyr"
author: "Joshua French"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_notebook: default
bibliography:
  - dwv.bib
  - packages_manip.bib
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE
)
knitr::write_bib(c("dplyr", "tibble", "tidyr", "palmerpenguins", "magrittr", "data.table", "reshape", "reshape2", "dbplyr", "dtplyr"), file = "packages_manip.bib")

```

This document is intended to help you quickly learn how to perform basic data frame manipulation using **dplyr** [@R-dplyr] and **tidyr** [@R-tidyr].

An accompanying YouTube playlist that walks through this document is available
by following the link [here](https://youtube.com/playlist?list=PLkrJrLs7xfbXaM7ukwxGu6xie3X64XVH3). I have also included direct, relevant video links throughout the file (e.g., immediately after the the relevant heading). An html version of this document can be downloaded to your current working direction (run `getwd()` in the Console to see where this is) by running the following command:
```{r, eval=FALSE}
download.file("https://raw.githubusercontent.com/jfrench/DataWrangleViz/master/04-data-frame-manipulation.nb.html",
              "04-data-frame-manipulation.nb.html")
```
The raw R Markdown code used to generate the html file can be obtained by toggling the "Code" box in the upper-right corner of the html file and selecting "Download Rmd".

# Data frame manipulation with **dplyr** and **tidyr** (Video: [YouTube](https://youtu.be/O5_Lj5FSX6E), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=a0c98d0f-f296-4184-ba69-af1d0126101e))

The **dplyr** and **tidyr** packages [@R-dplyr; @R-tidyr] are powerful packages for restructuring, enhancing, and manipulating data frames.

We'll discuss the **dplyr** package first, as its use cases are a bit simpler.

We load the **tibble**, **dplyr**, and **tidyr** packages for the discussion below. We also preemptively load the `penguins` data frame from the **palmerpenguins** package, as we use it heavily in our discussion.

```{r, message = FALSE, warning = FALSE}
library(tibble)
library(dplyr)
library(tidyr)
data(penguins, package = "palmerpenguins")
```

We will be looking at straightforward data frame manipulation examples. Depending on what you need to do, you may need to dig deeper into the packages we discuss or consider alternative packages. However, you should have a solid foundation for manipulating data frames.

Alternative data frame manipulation packages:

* **data.table** [@R-data.table]: almost a direct alternative to **dplyr** that manipulates `data.table` objects instead of tibbles (the enhanced **dplyr** data frame).
  * Faster than **dplyr**.
  * More memory efficient that **dplyr**.
  * Less accessible syntax?
* **dtplyr** [@R-dtplyr]: an implementation of **dplyr** that uses **data.table** internally.
  * The best of both worlds?
  * **dplyr** syntax is translated to **data.table** syntax.
  * Computation must be explicitly requested. 
* **dbplyr** [@R-dbplyr]: an implementation of **dplyr** that works with databases instead of tibbles.

Web searches about data frame manipulation will likely bring up references to the **reshape** [@R-reshape] and **reshape2** [@R-reshape2] packages.

* These packages are the spiritual predecessors to **tidyr**. 
* These packages do more than **tidyr**, but the syntax was more difficult.
* These packages aren't in active development, so it's better to avoid using them for active projects.
  * You may have a hard time getting help in using the older packages.

## Tibbles (Video: [YouTube](https://youtu.be/BlnebGyuOKM), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=3f16f13c-59de-4b21-83fb-af1d01260f69))
When the **dplyr** package interacts with a data frame, it usually produces a "tibble". 

* A tibble is supposed to be a "modern re-imagining" of R's standard data frame (referred to as `data.frame` for clarity).
* A tibble has class `tbl_df`.
* A tibble is also a `data.frame`, but has different default behaviors.
* Functions that take a `data.frame` should also take a tibble.
* Tibbles comes from the **tibble** package [@R-tibble], which is automatically imported by **dplyr**.

### Differences between a tibble and a `data.frame` (Video: [YouTube](https://youtu.be/_qHdqWx-vsQ), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=1066123d-3559-403a-92bb-af1d01260fa1))

A tibble generally has more restrictive behavior than a `data.frame`.

* Tibbles don't convert input types.
  * This isn't as big of a deal since the release of R 4.0.0 when the argument `stringsAsFactors` of the `data.frame` function was set to `FALSE` by default.
* Tibbles don't allow for partial matching of variable names.
* Tibbles doesn't add row names to the data frame.
* Subsetting a tibble always produces a tibble.
  * Subsetting a `data.frame` can return a vector, which can create edge-cases that results in a bug in your code!
* Tibbles do allow for non-valid R variable names, though I wouldn't do this intentionally.
  * A non-valid R variable name has to be surrounded by backticks. 
* Variables used to create a tibble must (generally) have the same length.
  * Unlike a `data.frame`, tibbles don't generally "recycle" the values of a vector when creating a data frame.
  * Tibbles only recycle vectors of length 1, i.e., it will recycle a vector of length 1 so that the new variable has length matching the other variables.
* The printing capabilities of a tibble are designed to be better than a `data.frame`. 
  * The user can directly specify the number of rows to print for a tibble.
  * Tibbles only print the number of columns that can fit horizontally in the Console.

### Example

Let's compare some of behaviors for a `data.frame` and a tibble. The last few lines use the `penguins` data frame from the **palmerpenguins** package [@R-palmerpenguins]. Run the following commands in the R Console.
```{r, eval=FALSE}
# create data.frame
dtf <- data.frame(numbers = 1:4, letters = letters[1:4], stringsAsFactors = TRUE)
# observe structure. look at the classes
str(dtf)
# partial matching of variable name
dtf$let
# subset single column of dtf
dtf[,c("letters")]
# subset single column of dtf no longer a data.frame!
is.data.frame(dtf[,c("letters")])
# add a new column to dtf
dtf$newv <- 1:2
# print dtf
dtf
# create tibble
tbl <- tibble(numbers = 1:4, letters = letters[1:4])
# observe structure. look at the classes
str(tbl)
# partial matching of variable name
tbl$let
# partial matching of variable name
tbl$letters
# subset single column of tbl
tbl[,c("letters")]
# subset single column of tbl still a tibble!
is_tibble(tbl[,c("letters")])
# add a new column to tbl of wrong length (not OK)
tbl$newv <- 1:2
# add a new column by recyclying a single value (OK)
tbl$newv <- 1
# add a new column with the same length as other columns (OK)
tbl$newv <- rep(1:2, times = 2)
# print tbl
tbl
# print only two rows of tbl
print(tbl, n = 2)
# print large, regular data.frame (penguins data)
as.data.frame(penguins)
# print large, regular tibble
penguins
```
### Subsetting/extracting parts of a tibble (Video: [YouTube](https://youtu.be/xR3x9Mbd_6A), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=ce3c36dd-32aa-477e-b715-af1d01260fdb))

The syntax used to subset or extract elements from a `data.frame` can be used for tibbles.

* As seen above, the result produced when using the syntax may differ for a `data.frame` and a tibble.

We will examine the behavior of a tibble when attempting to subset or extract its elements. We will use the `penguins` data set in the **palmerpenguins** package, which is a tibble. Execute the following commands in the R Console.
```{r, eval=FALSE}
# check whether the object is a tibble
is_tibble(penguins)
# extract the species column of penguins
penguins$species
# subset the first column of penguins
penguins[,1]
# subset the first and third columns of penguin
penguins[, c(1, 3)]
# subset the island variable of penguins
penguins[, c("island")]
# subset the island and bill_length_mm variables of penguins
penguins[, c("island", "bill_length_mm")]
# subset first 3 columns of penguins
penguins[1:3]
# subset the first 3 rows and named columns of penguins
penguins[1:3, c("island", "bill_length_mm")]
# subset all but the first 3 rows of penguins with the named columns 
penguins[-(1:3), c("island", "bill_length_mm")]
```

### Important functions for tibbles (Video: [YouTube](https://youtu.be/2zddbMKnxrY), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=031ea57c-4f34-4442-9612-af1d012626d0))

Below are some of the most important functions associated with tibbles:

* `tibble`: create a tibble similarly to how you would create a `data.frame`
* `as_tibble`: turn a `data.frame` into a tibble
* `is_tibble`: determine whether an object is a tibble
* `add_row`: add a row to a tibble
* `add_column`: add a column to a tibble

There are several other function associated with tibbles from the **tibble** package, but we don't discuss them here.

Run the following code in the R Console to see the results from applying the functions mentioned above.
```{r, eval = FALSE}
# create a simple tibble
tbl1 <- tibble(numbers = 1:3, letters = letters[1:3])
tbl1
# create a simple data.frame
dtf1 <- data.frame(numbers = 1:3, letters = letters[1:3])
dtf1
# convert the data.frame to a tibble
tbl2 <- as_tibble(dtf1)
# check that the objects are 
identical(tbl1, tbl2)
# confirm that tbl2 is a tibble
is_tibble(tbl2)
# add a row to the tibble
tbl1 <- add_row(tbl1, numbers = 4:5, letters = letters[4:5])
tbl1
# add a column to the tibble
tbl1 <- add_column(tbl1, cats = factor(rep(c("g1", "g2"), times = c(2, 3))), .after = "numbers")
tbl1
```
## Using **dplyr** (Video: [YouTube](https://youtu.be/kB-YSQ4r2Yc), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=ad40c79c-4c44-4d0b-9b4b-af1d01262826))

**dplyr** provides functions for the "verbs" (i.e., actions) related to common data frame manipulation tasks. The table below summarizes some of the most important functions by the aspect of a data frame it manipulates.

aspect | function | purpose
---|---|---
row | `filter` | selects rows based on a logical statement
row | `slice` | select rows based on position or other property
row | `arrange` | reorders the rows based on some property
row | `group_by` | groups a collection of rows hierarchically based on one or more variables
group of rows | `summarize` | produces a summarizing value or values for a group of rows
column | `select` | selects columns based on a logical statement, column names, etc.
column | `rename` | changes one or more column names
column | `mutate` | changes an existing column or adds a new column
column | `relocate` | reorders the columns

### The pipe operator

The pipe operator, `%>%`, is a popular approach for "piping" an object into the first argument (usually) of another function.

* The pipe operator was introduced by the **magrittr** package [@R-magrittr].
* It is extremely popular in the **tidyverse** for stringing sequences of operations together in a readable way.
* R version 4.1 will gain a native pipe operator (`|>`) that will be a better choice moving forward.

The pipe operator is best understood through example, as there are special cases that have to be considered.

### Filtering (Video: [YouTube](https://youtu.be/8tazJbgcDWg), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=598972a7-d218-4466-a453-af1d0126309a))

The `filter` function filters out the rows of a data frame that don't meet the designated criteria.

Let's `filter` the rows of the `penguins` data frame based on whether the `body_mass_g` variable is more than 6000. In the code below, the logical statement `body_mass_g > 6000` indicates the rows of `penguins` to retain. In the first example, we use the `filter` function to directly filter the `penguins` data frame. In the second example, we pipe the `penguins` data frame into the `filter` function and then use the logical statement to filter the rows.
```{r}
penguins[penguins$body_mass_g > 6000,]
penguins[penguins$body_mass_g > 6000 &
           !is.na(penguins$body_mass_g),]
filter(penguins, body_mass_g > 6000)
penguins %>% filter(body_mass_g > 6000)
```

### Slicing (Video: [YouTube](https://youtu.be/4_YptM2twH0), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=42da5bc1-17d0-42fe-bd30-af1d012638e9))

`slice` subsets rows of a data frame by index or with respect to other properties.

In the example below, we use the `slice` function to subset rows 10-12 of `penguins`, first directly, and then using the pipe operator.
```{r}
penguins[10:12, ]
slice(penguins, 10:12)
penguins %>% slice(10:12)
```

`slice_head` subsets the first `n` rows of a data frame. Conversely, `slice_tail` subsets the last `n` rows of a data frame.

* This is similar to the `head` and `tail` functions in **base** R.

In the examples below, we subset the first 2 rows, then the last two rows, of the `penguins` data frame.

```{r}
slice_head(penguins, n = 2)
penguins %>% slice_head(n = 2)
slice_tail(penguins, n = 2)
penguins %>% slice_tail(n = 2)
```

`slice_max` subsets the `n` rows of a data frame with the `n` largest values with respect to a specific variable. Conversely, `slice_min` subsets the `n` rows of a data frame with the `n` smallest values with respect to a specific variable.

In the examples below, we subset the rows of the `penguins` data frame with the largest and smallest `bill_lengh_mm` values, respectively. 

```{r}
slice_max(penguins, bill_length_mm, n = 2)
penguins %>% slice_max(bill_length_mm, n = 2)
slice_min(penguins, bill_length_mm, n = 2)
penguins %>% slice_min(bill_length_mm, n = 2)
```

`slice_sample` randomly selects `n` rows of a data frame. We randomly select 3 rows of `penguins` below. To make the results reproducible, we would need to set the random number seed using `set.seed` since the rows are randomly selected.

```{r}
slice_sample(penguins, n = 3)
penguins %>% slice_sample(n = 3)
```

### Arranging

The `arrange` function arranges the rows of a data frame with respect to a certain variable.

* By default, the rows are ordered in ascending order.
* The `desc` function can be used to arrange the rows in descending order.
* The ordering can be done within groups if specified.

In the code below, we arrange the rows of `penguins` in ascending and then descending order, respectively, with respect to `bill_length_mm`.

```{r}
arrange(penguins, bill_length_mm)
arrange(penguins, desc(bill_length_mm))
```
We could do the same thing using pipe operators using the code below.

```{r, eval = FALSE}
penguins %>% arrange(bill_length_mm)
penguins %>% arrange(desc(bill_length_mm))
```

The data frame can also be arranged within groups. In the code below, we select three rows from each `species` of `penguins` and assign it the name `penguins_slice`. We then  order the rows of `penguins_slice` by `bill_length_mm` in each group by setting `.by_group` to `TRUE`.

```{r}
penguins_slice <- penguins %>% group_by(species) %>% slice(1:3)
arrange(penguins_slice, bill_length_mm, .by_group = TRUE)
```

### Grouping and summarizing (Video: [YouTube](https://youtu.be/LaA7sVe76PA), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=dc18ab2d-2bd1-4333-b629-af1d01264000))

The `group_by` function groups rows of a data frame with respect to one or more variables, which can then be manipulated in various ways. 

`group_by` is often used in close connection to the `summarize` function, which computes statistical summaries for (grouped) rows into a single value.

Since we now want to string several operation in a row, the pipe operator is useful for stringing the commands together.

In the code below, we group the `penguins` data with respect to `species` and `island` and summarize the various aspects of the grouped data (compute the mean, the median, and the number of observations in each group for `bill_length_mm`.)

* We create variables with designated names inside `summarize`.
* `n()` counts the number of observations in each grouping.
* `na.rm` is need to compute the statistics without them being poisoned by `NA` values.

```{r, tidy = TRUE}
penguins %>%
  group_by(species, island) %>%
  summarize(mean_bill_length_mm = mean(bill_length_mm, na.rm = TRUE),
            median_bill_length_mm =  median(bill_length_mm, na.rm = TRUE),
            count = n())
```
### Selecting (Video: [YouTube](https://youtu.be/Su1VcLnIBuM), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=7334ec2d-56cc-4cfd-a5f8-af1d012647cc))

The `select` function can be used to select columns of a data frame. You can get pretty creative in how you select the columns (e.g., using string matching), but we only consider some simple examples.

* Run `?dplyr::select` for more complex examples.

We can select specific columns of a data frame by providing the column names. We select the `species` and `island` columns below of `penguins` below.

```{r}
select(penguins, species, island)
penguins %>% select(species, island)
```

Next, we select the columns of `penguins` that end in `mm`.

```{r}
select(penguins, ends_with("mm"))
penguins %>% select(ends_with("mm"))
```

### Column renaming (Video: [YouTube](https://youtu.be/5xuACH1a5UI), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=1f643690-5268-450d-beda-af1d01265367))

The `rename` function renames the columns of a data frame.

* The syntax is `new_name = old_name`.

In the example below, we rename the `bill_length_mm` column to `bill` and `flipper_length_mm` to `flipper`.

```{r}
# renaming columns using base R
names(penguins)
# names(penguins)[c(3, 5)] <- c("bill", "flipper")
# renaming columns using dplyr
rename(penguins, bill = bill_length_mm, flipper = flipper_length_mm)
penguins %>% rename(bill = bill_length_mm, flipper = flipper_length_mm)
```
### Mutating a column (Video: [YouTube](https://youtu.be/uiEzbFf5zJw), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=6ca2dc35-184d-4edb-be31-af1d01265e5c))

The `mutate` function modifies an existing column or can create a new column from the existing columns.

  * To create a new column, the syntax is `new_column = function(existing_columns)`
  * To modify an existing column, the syntax is `existing_column = function(existing_columns)`.
  
In the example below, we transform `bill_length_mm` into a new variable `bill_length_in` that describes the bill length in inches instead of millimeters (and then selects the two bill length columns)

```{r}
mutate(penguins, bill_length_in = bill_length_mm * 0.0393701) %>% 
  select(bill_length_mm, bill_length_in)
penguins %>% 
  mutate(bill_length_in = bill_length_mm * 0.0393701) %>%
  select(bill_length_mm, bill_length_in)
```

### Relocating a column (Video: [YouTube](https://youtu.be/efEvcF3VLMY), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=ffb59a39-649c-420c-81e9-af1d012a694b))

The `relocate` function moves one or more columns.

* By default the column is moved to the front of the data frame.
* The `.before` and `.after` arguments can be used to specify where the columns should be placed.

In the example below, we move the `sex` variable to the front of the data frame.

```{r}
relocate(penguins, sex)
penguins %>% relocate(sex)
```

In the next example, we relocate the `sex` and `year` columns to after the `species` column.

```{r}
relocate(penguins, sex, year, .after = species)
penguins %>% relocate(sex, year, .after = species)
```
In this example, we move all columns ending in `mm` to before the `island` column.

```{r}
relocate(penguins, ends_with("mm"), .before = island)
penguins %>% relocate(ends_with("mm"), .before = island)
```
## Tidy data (Video: [YouTube](https://youtu.be/uvDKLIK8i7E), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=9267676a-278c-4237-afe7-af1d012667cc))

@r4ds2017 and many others recommend that data scientists organize data frames so that they are "tidy". Tidy data [@r4ds2017; Chapter 12] is structured so that:

1. Each variable is in a single column and no variables share a column.
2. Each observation is in a single row.
3. Each data value has its own cell.

These principles are best explained with data sets that are not tidy. The **tidyr** package provides several illustrative data sets (`table1`, `table2`, `table3`, `table4a`, `table4b`, and `table5`) related to tidy data. These data sets provide data about tuberculosis cases for various countries in different years. The variables provided include:

* `country`: the name of the country
* `year`: the year an observation was observed
* `cases`: the number of tuberculosis cases observed in the specified year
* `population`: the population of the country in the specified year
* `rate`: the rate of disease observed in a country in a specified year (`cases/population`)
* `1999`, `2000`: values of a variable observed for a country in that year

### Columns

Each variable in a tidy data set should be in a single column and should not be in the same column as another variable.

`table2` in the **tidyr** package violates this rule. We print `table2` below.
```{r}
data(table2, package = "tidyr")
table2
```

Notice that the `cases` and `population` values are both in the `count` column. The values for `cases` and `population` should be in separate columns.

### Rows

Each observation in a tidy data set should be in its own row. 

Notice that `table2` also violates this principle. e.g., Rows 1 and 2 of `table2` are values for the same observation, Afghanistan in 1999.

### Values

Each value in a tidy data set should have its own cell. This means that you don't combine information about two variables into a single cell. This likely seem like an artificial use case, but non-data scientists are sometimes quite creative in how they enter their data!

`table3` in the **tidyr** package violates this rule. We print `table3` below.
```{r}
data(table3, package = "tidyr")
table3
```

Notice that the `rate` column is actually `cases/population`. It would not be a problem if this were a number (e.g., 745/19987071=0.0000372). The issue is that the values for `cases` and `population` are combined into a single column.

### Another non-tidy example

Times series data are often stored in a non-tidy way. In that case, the values of a variable are stored in a time-related column (like day or year). Different variables may then be stored across multiple data frames!

`table4a` and `table4b` in the **tidyr** package exhibit this type of non-tidyness. `table4a` provides the `cases` values for each `country` for each `year`. `table4b` provides the same thing for the `population` values.

```{r}
data(table4a, package = "tidyr")
data(table4b, package = "tidyr")
table4a
table4b
```

In these two tables, the values for `cases` and `population` are split across multiple columns. Specifically, the `cases` are split across columns by `year` (and similarly for `population`). Additionally, observations are not in a single row since the values of `cases` and `population` for each `country` for each `year` are not in a single row.

### A tidy example

A tidy version of the tuberculosis data is available in `table1`. This data frame satisfies the three requirements of tidy data.

```{r}
table1
```

## Using **tidyr** (Video: [YouTube](https://youtu.be/j5iH9ciMkL8), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=d5d79f4e-6aa6-4de9-b061-af1d01266a58))

The main functions in the **tidyr** package are `pivot_wider` and `pivot_longer`.

* `pivot_wider`: spreads a column into multiple columns. The resulting data frame will have fewer rows and more columns (wider).
* `pivot_longer`: gathers multiple columns into a single column. The resulting data frame will have more rows (longer) and fewer columns.

Tidying complicated data sets may require you to do seemingly odd steps like making a data frame longer and then widening some of the lengthened columns. Here are some tips for tidying a data frame:

1. Identify the ways in which your data are not tidy.
2. Identify a single aspect of the data that you want to make tidy (or at least closer to being tidy).
3. Tidy that single aspect of the data.
4. Repeat steps 2 and 3 until the data are tidy
5. If desired, combine the many tidying steps into a single command using the pipe operator.

### Widening data frames (Video: [YouTube](https://youtu.be/7cNxHl5aj-c), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=c01679b0-776d-4457-8a3c-af1d01267047))

The `pivot_wider` function increases the number of columns in a data frame but reduces the number of rows. The main arguments of the `pivot_wider` function are:

* `data`: the data frame to pivot
* `names_from`: the column from which the names of the new columns will be taken
* `values_from`: the column from which the values of the new columns will be taken

To tidy `table2`, we must widen it. We need to take the new column names from the `type` column and the values from the `count` column.

```{r}
table2
pivot_wider(table2, names_from = type, values_from = count)
table2 %>% pivot_wider(names_from = type, values_from = count)
```
### Lengthening data frames (Video: [YouTube](https://youtu.be/nhwcyk7bpQc), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=f632ce36-8528-4c02-a44d-af1d01268217))

The `pivot_longer` function decreases the number of columns in a data frame but increases the number of rows. The main arguments of the `pivot_longer` function are:

* `data`: the data frame to pivot
* `cols`: the columns from which to gather values
* `names_to`: the column name for the column that the names of the gathered columns are placed
* `values_to`: the name of the column to which the values are gathered

To tidy `table4a` and `table4b`, we must lengthen them. For `table4a` the new column name should be `cases`, while for `table4b` it should be `population`. The values for `cases` and `population` will come from the `1999` and `2000` columns. 

To tidy `table4a` and `table4b`, we need to gather the `1999` and `2000` columns into a single column and give them the appropriate variable name.

```{r}
table4a
tbl4a <- pivot_longer(table4a, c(`1999`, `2000`), names_to = "year", values_to = "cases")
tbl4a
table4b
tbl4b <- table4b %>% pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
tbl4b
```

### Merging/joining data frames (Video: [YouTube](https://youtu.be/-M4VIuemN08), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=9915b3cb-2c38-4cd8-9f8d-af1d0126a45e))

The remaining issue with our `tbl4a` and `tbl4b` data frames are that the information is spread across two data frames with some redundant information. We need to merge or join the two data frames into a single data frame.

The most streamlined solutions for this are to use the `base::merge` or `dplyr::*_join`functions to join the columns of the two data frames `by` their shared columns. By default, the `merge` function determines the identical columns in both data frames and joins the rows horizontally (removing redundant information). Alternatively, you can specify columns `by` which to join the rows.

We illustrate usage of the `merge` function below.

* The `x` and `y` arguments are the data frames we want to merge.
* The `by` argument indicates the shared columns we want to match the rows by.
* There are some other fancy arguments you can use if there are matching columns in the data frames that have different names.

Since `merge` is a **base** R function, the result is a `data.frame`.

```{r}
merge(x = tbl4a, y = tbl4b)
merge(x = tbl4a, y = tbl4b, by = c("country", "year"))
```

If we wish to stay within the **tidyverse** ecosystem, we can use one of the `*_join` functions in **dplyr**. These functions include `x`, `y`, and `by` arguments that work identically to those in the `merge` function.

* `left_join` merges the rows of `y` that match the rows of `x` into `x`.
* `right_join` merges the rows of `x` that match the rows of `y` into `y`.
* `inner_join` merges the rows of `x` and `y` that are in both data frames.
* `full_join` merges the rows of `x` and `y` that are in either data frame.
* See `?dplyr::inner_join` for examples

In this example, since rows of the two data frames match exactly, then any of the joins produce identical results. We will use `full_join`, since anecdotally, it is the most common type of join used.

```{r}
full_join(tbl4a, tbl4b)
```

## Separating and uniting columns (Video: [YouTube](https://youtu.be/GQMF2qNuZHI), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=30c7f0b2-a708-45c6-8a32-af1d0126b420))

The `rate` column of `table3` included both `cases` and `population`. We want to separate these values.

A column can be separated into multiple columns while retaining the same number of rows using the `separate` function. Important arguments include:

* `data`: the data frame to manipulate
* `col` the column to separate
* `into`: the names of the variables into which the column will be separated
* `sep`: the character or position by which to separate `col`
* `remove`: a logical value indicating whether `col` will be removed after separation. The default is `TRUE`.

```{r}
table3
separate(table3, col = rate, into = c("cases", "population"), sep = "/")
table3 %>% separate(col = rate, into = c("cases", "population"), sep = "/")
```

`table5` has the opposite problem of `table3`. It has columns `century` and `year` that shouldn't be split.

```{r}
table5
```

The `unite` function unites multiple columns of a data frame while retaining the same number of rows. Important arguments include:

* `data`: the data frame to manipulate
* `col` the name of the new column for the united values
* `...`: the columns to unite
* `sep`: the delimiter to separate the values in the columns
* `remove`: a logical value indicating whether the columns in `...` will be removed after unification. The default is `TRUE`.

We use the `unite` function to unite the `century` and `year` columns of `table5`.

```{r}
unite(table5, col = "year", "century", "year", sep = "")
table5 %>% unite(col = "year", "century", "year", sep = "")
```

For fun, can string together `unite` and `separate` operations to manipulate `table5` into a fully tidy data frame.

```{r}
table5 %>%
  unite(col = "year", "century", "year", sep = "") %>% 
  separate(col = rate, into = c("cases", "population"), sep = "/")
```
## References