---
pdf_document: default
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document: default
  pdf_document: default
  html_notebook: default
title: "Joshua French"
bibliography:
  - dwv.bib
  - packages_manip.bib
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE
)
knitr::write_bib(c("dplyr", "tibble", "tidyr", "palmerpenguins", "magrittr", "data.table", "reshape", "reshape2", "dbplyr", "dtplyr", "sf", "sp", "colorspace"), file = "packages_manip.bib")

```

# Wrangling and visualizing simple features spatial data in R

Spatial data can take many forms.

For a data scientist, spatial data is usually data for which each observation is geographic region or location.

> *Simple Features* (officially *Simple Feature Access*) is a set of standards that specify a common storage and access model of geographic feature made of mostly two-dimensional geometries (point, line, polygon, multi-point, multi-line, etc.) used by geographic information systems. It is formalized by both the Open Geospatial Consortium (OGC) and the International Organization for Standardization (ISO). [Wikipedia](https://en.wikipedia.org/wiki/Simple_Features)

Simple features are comprised of:

1. A geometry object (e.g., a point, a polygon, a line, etc.) 
2. Attribute data associated with the geometry object (e.g., the temperature at a certain time on a certain day, the number of new cases of a disease in a county in the last month).

## Simple features

As stated in the vignette to the **sf** package:

> Features have a geometry describing where on Earth the feature is located, and they have attributes, which describe other properties. The geometry of a tree can be the delineation of its crown, of its stem, or the point indicating its centre. Other properties may include its height, color, diameter at breast height at a particular date, and so on.

Just to clarify:

* The *geometry* of an observation describes where the object is located.
  * A geometry can be a point, a polygon (which is really just an ordered collection of points), or something more complicated.
* The *attributes* of a geometry object are what data scientists would usually consider the "data".

All geometries are made of points, which can be combined to create more and more complex objects.

A point can be 2-, 3-, or 4-dimensional.

The most common kinds of points are two-dimensional and are described using a 2-dimensional set of coordinates (X and Y), e.g., longitude and latitude or easting and northing.
  
  * 2-dimensional points are referred to as XY points.
  
A 3-dimensional point could include the X and Y coordinates as well as the altitude of the 2-dimensional point in 3-dimensional space.
  
  * The Z coordinate is another dimension denoting altitude of the point.
  * Combining a Z coordinate with an XY coordinate results in a 3-dimensional XYZ point.
  
Alternatively, a 3-dimensional point could include some other measure associated with the point.
  
  * The M coordinate is another dimension denoting some measure associated with the point.
  * It's pretty rare, but examples include time or measurement error.
  * Combining an M coordinate with an XY coordinate results in a 3-dimensional XYM point.
  
Combining X, Y, Z, and M coordinates results in a 4-dimensional point.

## Packages and tools for working with simple features

In what follows, we will use several packages.

* The **sf** package [@R-sf] is an R package for working with simple features (`sf` objects) both in terms of the geometry objects and the associated attributes.
  * The **sf** package can import, manipulate, and plot `sf` objects.
  * The **sf** package is intended to supersede the **sp** package [@R-sp], which is an older R package for working with spatial data.
  * Since the **sp** package is being superseded by the **sf** package, I recommend working with the **sf** package for spatial data analysis moving forward.
  
Because of how `sf` objects are represented in R, simple features (once constructed or imported) can be manipulated and plotted by many other well-known packages.

* The **dplyr** package [@R-dplyr] can be used to manipulate `sf` objects, and we will utilize it as certain times.
* The **ggplot2** package [@ggplot22016] can also be used to plot `sf` objects, which we will demonstrate.

Choosing a color palette to represent the attributes of your simple features is very important.

* The **grDevices** package included with **base** R provides  many color palettes.
  * The traditional palettes are `rainbow`, `heat.colors`, `terrain.colors`, `topo.colors`, and `cm.colors`.
  * The `hcl.colors` palette function was added in R 4.0.0 and provides numerous excellent color palettes.
  * Running the Examples in the documentation found in `?grDevices::hcl.colors` will show you examples of the available palettes through **grDevices**.
  * We provide color swatches for many palettes at the end of this tutorial (without the code).

* The **colorspace** package [@R-colorspace] provides many of the same HCL palettes available through the `hcl.colors` palette, but it also provides convenient functions for accessing this in **ggplot2**. 

We load all of these packages below, with the exception of **colorspace**, as it has a `coords` function that *masks* (i.e., replaces) a needed function in the **sf** package. So we will access the necessary **colorspace** function using the `::` approach.

```{r}
# library(colorspace)
library(sf)
library(dplyr)
library(ggplot2)
```

## Simple feature geometry objects
The most common simple feature geometry objects used by data scientists are:

geometry | function | description
--- | --- | ---
`POINT` |	`sf::st_point` | A geometry containing a single point
`POLYGON` |	`sf::st_polygon` | A geometry with a sequence of points that form a closed ring that doesn't intersect itself. Multiple rings form outer rings and holes within the polygon.

We go through the process of creating and plotting these geometries below.

```{r}
# create an XY point
(p1 = st_point(c(1,2)))
# create an XYZ point
(p2 = st_point(c(1,2,3)))
# the points look the same when plotted in two dimensions
plot(p1)
plot(p2)
# create a ring (connected set of points)
outer <- matrix(c(0, 0, 10, 0, 10, 10, 0, 10, 0, 0), ncol = 2, byrow = TRUE)
# create additional rings for holes
hole1 <- matrix(c(1, 1, 1, 2, 2, 2, 2, 1, 1, 1), ncol = 2, byrow = TRUE)
hole2 <- matrix(c(5, 5, 5, 6, 6, 6, 6, 5, 5, 5), ncol = 2, byrow = TRUE)
# combine rings into a list
pts <- list(outer, hole1, hole2)
# turn list of rings into a polygon
(pl1 <- st_polygon(pts))
# plot polygon
plot(pl1)
```

The other common geometry objects are:

geometry | function | description
--- | --- | ---
`LINESTRING` |	`sf::st_linestring` | A sequence of points that is connected with straight, non-self intersecting lines
`MULTIPOINT` |	`sf::st_multipoint` | A set of points
`MULTIPOLYGON` |	`sf::st_multipolygon` | A set of polygons
`MULTILINESTRING` |	`sf::st_multipoint` | A set of line strings
`GEOMETRYCOLLECTION` |	`sf::st_geometrycollection` | A set of the other geometries (except itself)

We provide examples of creating and plotting these geometries below.
```{r}
# create a matrix of multiple points
(pts <- matrix(rnorm(10), ncol = 2))
# convert matrix of points to linestring
(ls1 <- st_linestring(pts))
plot(ls1)
# convert matrix fo points to multipoints
(mp1 <- st_multipoint(pts))
plot(mp1)
# create multipolygons
pol1 <- list(outer, hole1, hole2)
pol2 <- list(outer + 24)
mp <- list(pol1, pol2)
(mpl1 <- st_multipolygon(mp))
plot(mpl1, axes = TRUE)
# create a multilinestring
(pts2 <- matrix(rnorm(6), ncol = 2))
(ml1 <- st_multilinestring(list(pts, pts2)))
plot(ml1)
# create a geometry collection
(gc <- st_geometrycollection(list(p1, pl1, ls1)))
plot(gc, col = "grey")
```

How do I need what type of geometry I need?

* Often, this is determined automatically when you read in shapefiles (which we'll discuss later).
* Attributes observed at a single location require a `POINT`.
* Most regions can be represented by a `POLYGON`.
* Complicated objects made of regions, e.g., an island chain like Hawaii, require `MULTIPOLYGONS`.
* The other types are for more complicated objects.
* There are 10 other rarer geometry types that we do not discuss (`CIRCULARSTRING`, `CURVE`, `SURFACE`, `TRIANGLE`, `COMPOUNDCURVE`, `CURVEPOLYGON`, `MULTICURVE`, `MULTISURFACE`, `POLYHEDRALSURFACE`, `TIN`). You can learn about them through the additional resources provided at the end of this document.

## Coordinate reference systems

A coordinate reference system (CRS) must be provided in order to place a point on the earth's surface.

When your import a geometry object from file, the CRS will often be provided.

* The WGS84 CRS is often the default for longitude/latitude coordinates.

Sometimes, in order to combine geometry objects, you must specify the CRS of a geometry object.

There are many CRSs. A CRS is often used because it is known to have a certain desirable property. A discussion of CRSs is beyond the scope of this tutorial. And when you do need to know about CRSs, it will probably be so specific that a general discussion won't help. However, here are a few references related to CRSs that may provide a bit more detail:

* [QGIS CRSs](https://docs.qgis.org/3.16/en/docs/gentle_gis_introduction/coordinate_reference_systems.html)
* [Introduction to CRSs](https://bookdown.org/tep/gisbooklet/introduction-to-coordinate-reference-systems.html)

## `sf` objects

An `sf` object is a `data.frame` that has a simple feature geometry list column (i.e., a column of geometry objects). So you can work with `sf` objects similar to how you would work with a `data.frame` object, though it may have different default behaviors because it has been extended to an `sf` object.

* The geometry-list column contains the geometry object for each observation.
* The other columns of the `sf` object contain the attributes of the geometry object.
* Each row of the `sf` object is a simple feature. Alternatively, each observation is a simple feature. 

In practice, `sf` objects are often initially created by reading in a shapefile. However, particularly for `POINT` observations, you may need to create an `sf` object manually.

In what follows, we create `sf` objects for `POINT` geometry objects and `POLYGON` geometry objects.

* We can use the same previously discussed functions (e.g., `sf::st_point`, `sf::st_polygon`, etc.) to create the geometry objects. 
* The `sf::st_sfc` function is used to create a geometry list column.
* The `sf::st_sf` function is used to combined a `data.frame` of attributes with the geometry list column.

```{r}
# create POINT objects
pt1 <- st_point(c(0, 0))
pt2 <- st_point(c(0, 1))
pt3 <- st_point(c(1, 1))
# create geometry list column
glc1 <- st_sfc(list(pt1, pt2, pt3))
# is glc1 a list?
is.list(glc1)
# what class is glc1
class(glc1)
# create attribute data.frame with temperature and rainfall attributes
df1 <- data.frame(temperature = c(10, 11, 10.4), rainfall = c(1, 1.3, 0.9))
# create sf object
sf1 <- st_sf(df1, geometry = glc1)
# plot sf1
plot(sf1["temperature"], pch = 20, axes = TRUE)
# create polygon objects
outer1 <- matrix(c(0, 0, 1, 1, 0, 0, 1, 1, 0, 0), ncol = 2)
pl1 <- st_polygon(list(outer1))
# outer2 is outer1 shifted 1 unit to the right
outer2 <- matrix(c(1, 1, 2, 2, 1, 0, 1, 1, 0, 0), ncol = 2)
pl2 <- st_polygon(list(outer2))
# create geometry list columns
glc2 <- st_sfc(list(pl1, pl2))
# what class is glc2
class(glc2)
# create second sf object (only include an attribute column and geometry)
sf2 <- st_sf(cases = c(10, 12), geometry = glc2)
# plot sf2
plot(sf2)
# what happens if you combine geometry types?
glc3 <- st_sfc(list(pt1, pt2, pt3, pl1, pl2))
# what class is glc3
class(glc3)
# create sf3
sf3 <- st_sf(attribute1 = rnorm(5), geometry = glc3)
# plot sf3
plot(sf3, pch = 20, pal = topo.colors)
# plot sf3 with even breaks
plot(sf3, pch = 20, pal = topo.colors, breaks = seq(min(sf3$attribute1), max(sf3$attribute1), length = 12))
```

## Importing shapefiles

A data scientist is most likely to work with `sf` objects obtained from importing a shapefile into R. 

ArcGIS defines a shapefile as:

> A shapefile is a simple, nontopological format for storing the geometric location and attribute information of geographic features. Geographic features in a shapefile can be represented by points, lines, or polygons (areas). [What is a shapefile?](https://desktop.arcgis.com/en/arcmap/10.3/manage-data/shapefiles/what-is-a-shapefile.htm)

Generally, a shapefile can be imported into R as an `sf` object.
  
  * Each row is an observation related to a `geometry` object.
  * Most of the columns will represent a different attribute of the `geometry` object.
  * There will be a `geometry` column that contains the `geometry` object for each observation.
  
Shapefiles are widely available for describing many different spatial objects like counties, census tracts, zip code tabulation areas (ZCTAs), states, countries, etc. There are even shapefiles that can be used to describe other spatial objects like roads, rivers, lakes, etc.

  * A simple web search with appropriate terms will usually bring up a website with a relevant shapefile for your data.
  * e.g., search "usa shapefile" brings up a [Census bureau page](https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html) with many different shapefiles for different areas and characteristics of the United States. 
  * A "shapefile" is often a zipped folder that contains many files inside it.
  * The `.shp` file is usually the file you want to import.
  
We can import that shapefile into R as an `sf` object using the `st_read` function.
  * Typically, we want to provide the `shp` file to the `dsn` argument (data source name) of `st_read`.

In this case, I have downloaded the `cb_2018_us_state_20m.zip` containing a shapefile of the U.S. states. I have unzipped this file into the `cb_2018_us_state_20m` folder in the `data` folder in current working directory.

  * The current working directory is location to which  R automatically reads or saves files.
  * The can learn what your current working directory by running `getwd()` in the Console.
  * You can change your current working directory by running `setwd(path)` in the Console, where `path` is the directory you want to set as your working directory.
  
In our case, we want to read the file `cb_2018_us_state_20m.shp` in the `cb_2018_us_state_20m` folder in the `data` folder of our current working directory. We can run the following command to import the desired shapefile.
  * "`./`" indicates the current directory.
  * The `file` argument "`./data/cb_2018_us_state_20m/cb_2018_us_state_20m.shp` tells R to look for the `cb_state_20m.shp` file in the file path `./data/cb_state_2018_us_state_20m`.

For simplicity, I have uploaded `cb_2018_us_state_20m.zip` to my GitHub repository. Prior to importing the the `.shp` file, running the `download` and `unzip` commands should download and unzip the necessary folder into your working directory so that can run the same code to import the shapefile.

```{r}
# download.file("https://github.com/jfrench/DataWrangleViz/raw/master/data/cb_2018_us_state_20m.zip", destfile = "./data/cb_2018_us_state_20m.zip", method = "auto")
# unzip("./data/cb_2018_us_state_20m.zip", exdir = "./data/cb_2018_us_state_20m")
us_sf <- st_read(dsn = "./data/cb_2018_us_state_20m/cb_2018_us_state_20m.shp")
```

The `st_read` provides helpful information automatically when run. In this case, we learn:

  * We imported an `sf` object with 52 features (observations).
  * The imported `sf` object has 9 attributes (columns).
  * The features appear to be the simple feature geometry `MULTIPOLGYON`.
  * The dimension is `XY`, so we are working with 2-dimensional data.
  * The bounding box tells us the largest and smallest y-coordinates.
  * The CRS is NAD83.

We use the `class` function to see classes of `us_sf`.
```{r}
class(us_sf)
```

`us_sf` is a `data.frame` that has been extended to the `sf` class.

Let's use the `str` function to learn more about the structure of `us_sf`.
```{r}
str(us_sf)
```

`us_sf` has 52 rows and 10 columns. The (most) useful columns are:

  * `STUSPS`: the state abbreviation
  * `NAME`: the state name
  * `ALAND`: the land area of each state
  * `AWATER`: the water area of each state
  
The `geometry` column is the simple feature geometry list column and contains the  `geometry` object for each observation.

# Wrangling `sf` objects

An `sf` object is a type of data frame, similar to how a tibble is a type of data frame. Both classes extend the `data.frame` class.

  * This means that you can work with these objects similar to how you would work with a `data.frame`, though the default behaviors may be different.
  * We can use the **dplyr** package or similar to manipulate an `sf` object.
  
We can select column of `us_sf`

```{r}
us_sf$STUSPS
us_sf[,5]
us_sf[,"STUSPS"]
us_sf["STUSPS"]
```
Note that the `$` operator extracts the column from `us_sf` (returning a `character` vector), while the other choices *subset* `us_sf` and remain `sf` objects (i.e., the geometry list column is retained).

```{r}
us_sf$STUSPS
us_sf[,5]
us_sf[,"STUSPS"]
us_sf["STUSPS"]
```

We can select rows in a similar fashion.

```{r}
library(dplyr)
us_sf[2:3,]
us_sf[us_sf$STUSPS == "CO",]
us_sf[startsWith(us_sf$STUSPS, "C"),]
us_sf %>% filter(startsWith(us_sf$STUSPS, "C"))
```

Note that `startsWith` is a **base** R function that finds the elements of a `character` vector that start with a certain set of characters while `start_with` is a **dplyr** function that is used to select columns of a data frame based on a pattern.

us_sf[CO,]

What is really cool is that we can use a spatial object as a row selector. Let's extract the  "Colorado" row from `us_sf`.

```{r}
co <- us_sf[us_sf$NAME == "Colorado",]
```

If we pass the `co` object as the row argument inside the square brackets, `[]`, then the rows of `us_sf` with `geometry` objects that intersect the `co` `geometry` object will be returned

```{r}
(co_intersects <- us_sf[co,])
plot(co_intersects["NAME"])
plot(st_geometry(co_intersects))
```
```{r}
caco <- us_sf[us_sf$NAME == "Colorado" | 
                us_sf$NAME == "California",]
(caco_intersects <- us_sf[caco,])
plot(st_geometry(caco_intersects))
```
The **base** `merge` and **dplyr** `*_join` functions can be used to merge a data frame with an `sf` object.

  * The `sf` object must be the first argument of these functions.
  
Let's access a COVID-19 related data frame available in the **bayesutils** package, which can be installed from GitHub. 

* We install the package from GitHub.
* We then load the `covid_20210307` dataset from the **bayesutils** package.

```{r}
# remotes::install_github("jfrench/bayesutils")
data("covid_20210307", package = "bayesutils")
```

The `state_abb` column of `covid_20210307` has the state abbreviations and matches the `STUSPS` column of `us_sf`.

We use the **base** `merge` function to unite these two objects into a new object, `covid_us`.

```{r}
covid_us <- merge(us_sf, covid_20210307,
                  by.x = "STUSPS", by.y = "state_abb")
head(covid_us)
```
Alternatively, we could have used a **dplyr** `*_join` function. We'll use `full_join`. Note the special syntax in the `by` argument to address the fact that we want to merge the rows based on different columns in the data frames.
```{r}
covid_us_tbl <- full_join(us_sf, covid_20210307,
                          by = c("STUSPS" = "state_abb"))
print(covid_us_tbl, 5)
```

If a new row is added to the `sf` object without a corresponding `geometry`, then an empty `geometry` is added for that row.

## Plotting `sf` objects

The power of the `sf` package is the ability to easily create plots of spatial data.

### Plotting `sfc` objects


To simply plot the list-column of `geometry` objects of an `sf` object, you can use:

* `st_geometry` to extract the list-column of simple feature geometries from the `sf` object (this will be an object of class `sfc`).
* `plot` to plot the `sfc` object.

```{r}
plot(st_geometry(us_sf))
```
Alternatively, you can directly extract the `sfc` component of the `sf` object using `$geometry`, then plot the `sfc` object.

```{r}
plot(us_sf$geometry)
```

You can easily change aspects of the plotted `sfc` object (or an `sf`) object using the standard arguments:

* `axes` can be set to `TRUE` to show the axes
* `xlab` and `ylab` will change the axis labels
* `xlim` and `ylim` can be used to constrain the plotting regions.
  * Note that "W" longitude are indicated using negative numbers, while "E" longitudes are positive numbers.
  * Note the latitude values in the northern hemisphere are positive numbers. Latitude values in the southern hemisphere are negative numbers.

Each geometry has specific arguments that the user can change. In this case, we can change the fill color of the `MULTIPOLYGON` objects using the `color` argument and the border using the border argument.

```{r}
plot(us_sf$geometry, axes = TRUE,
     xlab = "longitude", ylab = "latitude",
     col = "grey", border = "blue",
     xlim = c(-125, -75),
     ylim = c(22, 50))
```
You can change the colors of the individual simple features with a little creativity. Let color all the "C states" (California, Colorado, Connecticut) a little differently.

```{r}
# determine indices of C states
c_states <- startsWith(us_sf$STUSPS, "C")
# create a character vector of replicated "grey"
# values matching the number of rows in us_sf
mycol <- rep("grey", nrow(us_sf))
#change "grey" to "orange" for the c_states indices
mycol[c_states] <- "orange"
# create a character vector of replicated "yellow"
# values matching the number of rows in us_sf
myborder <- rep("yellow", nrow(us_sf))
#change "yellow" to "blue" for the c_states indices
myborder[c_states] <- "blue"
# create the customized plot
plot(st_geometry(us_sf), col = mycol, border = myborder, xlim = c(-125, -75),
     ylim = c(22, 50))
```

### Plotting attributes of an `sf` objection

By default, use you can use the `plot` function to plot all the attributes of an `sf` object. In general, this isn't very useful.

```{r}
plot(covid_us)
```

More likely, you will want to plot a single attribute (variable), which can be done by subsetting that variable in the `sf` object and plotting the subsetted object.

Let's plot the land area of the states, excluding Alaska and Hawaii. First, we exclude the Alaska and Hawaii rows (and save the filtered object).

```{r}
covid_us <- covid_us %>% filter(!is.element(STUSPS, c("AK", "HI")))
plot(covid_us["ALAND"])
```

Land area is directly related to the area of the state. It would be interesting to visualize the states that have the greatest percentage of water. Let's create a new variable in `covid_us` that computes the percentage of the state that is water. We'll then plot this variable (excluding Alaska and Hawaii)

```{r}
covid_us <- covid_us %>% mutate(prop_water = AWATER/(AWATER + ALAND))
plot(covid_us["prop_water"])
```

Not surprisingly, coastal states and states on the Great Lakes have the highest percentage of water.

We can use the following commands to identify the states with the 6 largest proportions of water.
```{r}
covid_us %>% slice_max(prop_water, n = 6) %>% select(NAME, prop_water)
```

We can change the number of breaks in our color bar via the `nbreaks`.

We can change the colors in our color bar by specifying the `pal` argument.

* The `pal` argument takes a function that, when given `n`, the number of desired colors, returns a vector of colors.
* The `hcl.colors`, `rainbow`, `heat.colors`, `terrain.colors`, `topo.colors`, and `cm.colors` are all color palettes in base R that can be used to change the colors palettes.
* The `hcl.colors` palette is particularly good, as it includes color palettes `viridis` and `cividis` (corrected viridis) that are particularly well-suited to displaying colors that are color-blind and can be understood when printed in black and white.
* The **colorspace** package also has a wide variety of palettes you might consider.

However, the `hcl.colors` function has a `palette` argument to specify the desired palette. To use an `hcl.color` palette when plotting an `sf` object, we need to build a custom palette that only requires the number of colors desired. We create those function for the viridis and cividis palettes below. We then see that it produces the desired results.

```{r}
viridis_pal <- function(n) hcl.colors(n, palette = "viridis")
cividis_pal <- function(n) hcl.colors(n, palette = "cividis")
# dirty approach to see colors in the palette
image(matrix(0:4), col = viridis_pal(5))
image(matrix(0:4), col = cividis_pal(5))
```


Let's do some analysis of the actual COVID-19 data. Let's create a new column for `death_rate_100k`, which is the number of confirmed and probable COVID-19 deaths per 100,000 persons. Let's display the death rate using the viridis palette and then the cividis palette.

```{r}
covid_us <- covid_us %>% mutate(death_rate_100k = deaths/population*100000)
plot(covid_us["death_rate_100k"],
     nbreaks = 5,
     pal = viridis_pal)
plot(covid_us["death_rate_100k"],
     nbreaks = 5,
     pal = cividis_pal)
```

### Plotting `sf` objects with **ggplot2**

The **ggplot2** package includes a geometry for `sf` objects, `geom_sf`, the is typically adequate for using **ggplot2** to produce graphics for `sf` objects.

* The `sf` object is passed as the `data` argument to the `ggplot` function.
* `geom_sf` is the geometry of the `data`
* The `XY` coordinates of the `sf` object are automatically mapped to `x` and `y` aesthetics.
* The `color`, `linetype`, `fill`, etc., of the geometry can be changed by mapping the an attribute of the `sf` object to the appropriate aesthetic.

We will create a choropleth map of the `covid_us` data.
  * A choropleth map is a map of regions colored to indicate the level of some variable associated with the regions.

```{r}
ggplot(covid_us) + geom_sf(aes(fill = death_rate_100k))
```

```{r}
ggplot(covid_us) + geom_sf(aes(fill = death_rate_100k)) +
  colorspace::scale_fill_continuous_sequential(palette = "viridis")
```

## Additional resources

**sf** help and tutorials from the package authors [https://r-spatial.github.io/sf/]

*Geocomputation with R* by Robin Lovelace [https://bookdown.org/robinlovelace/geocompr/]. This book covers tons of aspect of spatial data (not just from an R perspective, but general theory and concepts). This will help you to learn a lot more about representing and working with spatial data in general, not just working with it in R.

*Spatial Data Science with applications in R* by Edzer Pebesma and Roger Bivand [https://keen-swartz-3146c4.netlify.app/]. Also covers much much about representing spatial data than you probably ever thought possible! The authors are the main creators of the **sf** package.

## Color palettes available with **base R**

```{r, echo = FALSE}
require("graphics")

par(mfrow = c(1, 1))

## color swatches for RGB/HSV palettes
demo.pal <-
  function(n, border = if (n < 32) "light gray" else NA,
           main = paste("color palettes;  n=", n),
           ch.col = c("rainbow(n, start=.7, end=.1)", 
                      "heat.colors(n)",
                      "terrain.colors(n)", "topo.colors(n)",
                      "cm.colors(n)")) {
    nt <- length(ch.col)
    i <- 1:n; j <- n / nt; d <- j/6; dy <- 2*d
    plot(i, i+d, type = "n", yaxt = "n", ylab = "", main = main)
    for (k in 1:nt) {
        rect(i-.5, (k-1)*j+ dy, i+.4, k*j,
             col = eval(str2lang(ch.col[k])), border = border)
        text(2*j,  k * j + dy/4, ch.col[k])
    }
}
demo.pal(16)

## color swatches for HCL palettes
hcl.swatch <- function(type = NULL, n = 5, nrow = 4,
  border = if (n < 15) "black" else NA) {
    palette <- hcl.pals(type)
    cols <- sapply(palette, hcl.colors, n = n)
    ncol <- ncol(cols)
    nswatch <- min(ncol, nrow)

    par(mar = rep(0.1, 4),
        mfrow = c(1, min(5, ceiling(ncol/nrow))),
        pin = c(1, 0.5 * nswatch),
        cex = 0.7)

    while (length(palette)) {
        subset <- 1:min(nrow, ncol(cols))
        plot.new()
        plot.window(c(0, n), c(0, nrow + 1))
        text(0, rev(subset) + 0.1, palette[subset], adj = c(0, 0))
        y <- rep(subset, each = n)
        rect(rep(0:(n-1), n), rev(y), rep(1:n, n), rev(y) - 0.5,
             col = cols[, subset], border = border)
        palette <- palette[-subset]
        cols <- cols[, -subset, drop = FALSE]
    }

    par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1), cex = 1)
}

hcl.swatch("qualitative")
hcl.swatch("sequential")
hcl.swatch("diverging")
hcl.swatch("divergingx")
```

## References
