---
title: "Interactive R graphics with plotly"
author: "Joshua French"
output:
  html_document: default
bibliography:
- dwv.bib
- packages_interactive.bib
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```

```{r, include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'tidyverse', 'bookdown', 'knitr', 'rmarkdown', 'ggplot2', 'dplyr', 'purrr', 'tidyr', 'readr', 'stringr', 'tibble', 'forcats', 'palmerpenguins', 'plotly', 'ggiraph', 'shiny', 'htmlwidgets', 'MASS', 'sf', 'rgl'), 'packages_interactive.bib')
```

# Best R packages for creating interactive graphics (Video: [YouTube](https://youtu.be/cdrUfPAK63I), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=ea6b49dc-aad0-48e6-9278-af430161ecd0))

Interactive graphics are visual displays that dynamically provide information to users based on the user interacting with the graphic.

The best packages for creating interactive graphics in R are:

- **plotly** [@R-plotly]: provides functions to make **ggplot2** graphics interactive and a custom interface to the JavaScript library **plotly.js** inspired by the grammar of graphics. This is perhaps the best known interactive visualization library.
- **ggiraph** [@R-ggiraph]: creates interactive **ggplot2** graphics using **htmlwidgets** [@R-htmlwidgets].
- **rgl** [@R-rgl]: provides functions for 3D interactive graphics using OpenGL or to various standard 3D file formats 
- **shiny** [@R-shiny]: a package for creating interactive web apps.

We use **plotly** to create a surface plot of the Maunga Whau volcano data.

```{r, message=FALSE, warning = FALSE, echo = FALSE}
library(plotly, quietly = TRUE)
plot_ly(z = ~volcano, type = "surface")
```

We use **ggiraph** to provide an example of an interactive graphic using the `starwars` data from the **dplyr** package [@R-dplyr].

```{r, echo = FALSE}
# fit the robust linear model
library(ggplot2)
library(ggiraph)
data(starwars, package = "dplyr")
rlmod <- MASS::rlm(mass ~  height, data = starwars)
# extract the coefficients
coeffs <- rlmod$coefficients
# turn coefficients into a useful character string
equation <- paste0("intercept: ", round(coeffs[[1]], 2),
                   "\nslope: ", round(coeffs[[2]], 2))
iscatter5 <- ggplot(data = starwars,
                    aes(x = height, y = mass)) + 
  geom_point_interactive(aes(data_id = name,
                             tooltip = name)) +
  geom_abline_interactive(intercept = coeffs[1],
                      slope = coeffs[2],
                      tooltip = equation,
                      data_id = coeffs[1],
                      color = "blue")
girafe(ggobj = iscatter5)
```

We use **rgl** to display 3-dimensional perspective plot with contour levels of the Maunga Whau volcano.

```{r, echo = TRUE, message = FALSE, warning = FALSE, webgl = TRUE, eval = TRUE}
library(rgl)
z <- 2 * volcano        # Exaggerate the relief
x <- 10 * (1:nrow(z))   # 10 meter spacing (S to N)
y <- 10 * (1:ncol(z))   # 10 meter spacing (E to W)

open3d()
id <- persp3d(x, y, z, aspect = "iso",
      axes = FALSE, box = FALSE, polygon_offset = 1)
contourLines3d(id)     # "z" is the default function
filledContour3d(id, polygon_offset = 1, nlevels = 10, replace = TRUE)
```

A Shiny app example related to NCAA swim teams can be found at [https://shiny.rstudio.com/gallery/ncaa-swim-team-finder.html](https://shiny.rstudio.com/gallery/ncaa-swim-team-finder.html).

# Two approaches with **plotly** for creating interactive graphics (bar plot) (Video: [YouTube](https://youtu.be/Pt5Whpx_COM), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=32c3fa9c-7556-49f7-b68e-af43017e7a63))

The easiest way to create interactive graphics is to:

1. Create a plot using **ggplot2**.
2. Use the `ggplotly` function from the **plotly** package to make the graphic interactive.

The advantage of this approach is that it is dead simple. 

The disadvantage of this approach is that it may not give
us the desired control over the aspects of the graphic that are interactive. We produce numerous interactive graphics using this approach below.

As stated on the plotly website ([https://plotly.com/r/getting-started/](https://plotly.com/r/getting-started/)):

> plotly is an R package for creating interactive web-based graphs via the open source JavaScript graphing library plotly.js.

It can be used to add interactivity to plots created with **ggplot2** or create interactive plots on its own.

First, we load the necessary packages. We load the **ggplot2** and **plotly** packages to create the graphics and load the `penguins` data set from the **palmerpenguins** package to load the data we will plot.

```{r, message=FALSE, warning=FALSE}
library(ggplot2, quietly = TRUE)
library(plotly, quietly = TRUE)
data(penguins, package = "palmerpenguins")
```

In the code below, we use **ggplot2** to create a basic bar plot of penguin `species`. We assign this graphic the name `ggbar`.

We then use the `ggplotly` function in the **plotly** package to make the graphic interactive.

The interactive graphic provides the frequency associated with each species when we hover over a bar.

```{r, warning = FALSE}
# bar plot of penguin species
ggbar <-
  ggplot(penguins) +
  geom_bar(aes(x = species))
# make bar plot interactive
ggplotly(ggbar)
```

Next, we use the direct capabilities of the **plotly** package to create a bar plot of penguin `species`. In general, the `plot_ly` function in the **plotly** package is all we need to create basic interactive graphics. We can also add additional layers to the graphics using various `add_*` functions and customize the layout using the `layout` functions. 

The main arguments to the `plot_ly` function are:

- `data`: an optional data frame whose variables will be plotted. To access a variable in `data`, we must use `~` before the variable's name.
- `type`: a character string indicating the type of plot to create, e.g., `"bar"`, `"histogram"`, `"box"`, `"violin"`, `"scatter"`, 
- `...`: arguments passed to the plot type that specify the *attributes* of the graphic (which is similar to the aesthetics in **ggplot2**), e.g., `x`, `y`, etc.
- `split`: Discrete values used to create multiple traces (one trace per value). This is similar to the `group` argument in **ggplot2**. A "trace" describes "a single series of data in a graph" ([https://plotly.com/r/reference/index/](https://plotly.com/r/reference/index/)).
- `color`: values mapped to a fill color.
- `alpha`: a number between 0 and 1 controlling the transparency of the graphic.

To create a bar plot using **plotly**, we need a data frame containing the count associated with each level of the categorical variable we want to display. 

We first create a data frame that summarizes the counts of each `species`. We use the `group_by`, `summarize`, and `n` functions from **dplyr** [@R-dplyr] to do this.

```{r, warning = FALSE}
# create data frame of frequency for each species
species_counts <-
  penguins |>
  dplyr::group_by(species) |>
  dplyr::summarize(frequency = dplyr::n())
# print data frame
print(species_counts, n = 3)
```

Once we have a data frame that describes the frequency associated with each level of the categorical variable, we can create a bar plot using **plotly**. In the `plot_ly` function, we: 

- Set the `type` argument to `bar`
- Associate the levels of the categorical variable with the `x` attribute.
- Associate the frequency of each level with the `y` attribute.

The interactive graphic provides the frequency associated with each species when we hover over a bar. 

```{r, warning = FALSE}
# create interactive bar chart
plot_ly(species_counts,
        x = ~species,
        y = ~frequency,
        type = "bar")
```

# Interactive histograms (Video: [YouTube](https://youtu.be/4ic0HOws-Dg), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=2302d1d2-5d76-48e6-be47-af46011aab64))

We now create interactive histograms using two approaches.

In the code below, we:

1. Use **ggplot2** to create a basic histogram of the `bill_length_mm` variable for the `penguins` data. We assign this graphic the name `gghist`.
1. Use `plotly::ggplotly` to make the graphic interactive.

The interactive graphic indicates the midpoint of each bin and the number of penguins falling in each bin.

```{r, warning = FALSE}
gghist <-
  ggplot(penguins) +
  geom_histogram(aes(x = bill_length_mm))
ggplotly(gghist)
```

To create a similar histogram using the `plot_ly` function, we:

- Set the `type` arugment to `histogram`.
- Associate `bill_length_mm` with the `x` attribute.
- Set the`nbinsx` argument to control the number of bins in the histogram.

The interactive histogram indicates the endpoints of each bin and the number of penguins in each bin.

```{r, warning = FALSE}
plot_ly(penguins,
        x = ~bill_length_mm,
        type = "histogram",
        nbinsx = 30)
```

The histogram produced by the `plot_ly` function looks a bit different from the histogram produced by **ggplot2** because the locations of the bins are different. To make them the safe, we can use the `ggplot2::layer_data` to get the "under the hood" information **ggplot2** uses to produce its plot.

In the code below, we use `layer_data` to access the internal data used by **ggplot2** to create a histogram. The `xmin` variable indicates the lower bound of each histogram bin. The lower bound of the far left bin starts at 31.72724. We then use the `diff` function to determine the bin width (this computes the difference between success lower bounds). 

```{r, warning = FALSE}
# get histogram data from gghist
datahist <- layer_data(gghist)
# determine starting point
head(datahist$xmin, 3)
# determine bin width
head(diff(datahist$xmin), 3)
```

Now that we know the start location of the left most bin and the size (width) of the bins, we can pass these arguments as `start` and `size` arguments to a named list for the `xbins` argument to `plot_ly`. This will create an interactive histogram that mimics the one produced by **ggplot2**.

```{r, warning = FALSE}
plot_ly(penguins,
        x = ~bill_length_mm,
        type = "histogram",
        xbins = list(start = 31.76724,
                     size = 0.9482759))
```

# Interactive density plot (Video: [YouTube](https://youtu.be/B_xaEQ5B4Tc), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=fcfeaf24-2cdf-4d62-872e-af460142c423))

We examine how to construct interactive density plots using two approaches.

In the code below, we:

1. Use **ggplot2** to create a density plot of `bill_length_mm` for each `species` that uses semi-transparent color to distinguish the different `species`. We assign this plot the name `ggdens`.
1. Use `plotly::ggplotly` to make the graphic interactive.

The interactive graphic indicates the `species`, `bill_length_mm`, and density when we hover over a density curve.

```{r, warning = FALSE}
ggdens <-
  ggplot(penguins) +
  geom_density(aes(x = bill_length_mm, fill = species), alpha = 0.3)
ggplotly(ggdens)
```

Surprisingly, there is no easy way to create a standard density plot natively using `plot_ly`.

To work around this, we can manually create the density information using the `base::density` function and then extract the associated `x` and `y` of the density information. However, we want to do this individually for each species, so we instead use the `ggplot2::layer_data` function to get the same information from our previous **ggplot2** graphic. 

We assign the name `dens_data` to the information from the `layer_data` function. `dens_data` stores the relevant density curve information in the `x` and `density` variables, while the `group` variable distinguishes the different `species`. We turn the `group` variable into a factor with the correct `species` names.

```{r, warning = FALSE}
# extract density data from ggdens
dens_data <- layer_data(ggdens)
# view data
head(dens_data, n = 3)
# convert group to factor
dens_data$group <-
  factor(dens_data$group,
         labels = c("adelie", "chinstrap", "gentoo"))
```

We want to trace the density curves for each species in a scatter plot that connects the poitns for each species. Using `plot_ly`, and the `dens_data` data frame, we:

- Associate the `x` variable with the `x` attribute.
- Associate the `density` variable with the `y` attribute.
- Associate the `group` variable with the `split` argument. This is roughly equivalent to the `group` aesthetic in **ggplot2**.
- Specify `type = "scatter"` to produce a scatter plot.
- Specify `mode = "line"` to connect the points using a line but not show the points (markers) themselves.
- Combine the native R pipe, `|>` with the `layout` function to change the x-axis label.

The resulting interactive density plot indicates the value of `bill_length_mm` for each species and the associated `density`. 

```{r, warning = FALSE}
plot_ly(dens_data,
        x = ~x,
        y = ~density,
        split = ~group,
        type = "scatter",
        mode = "line") |>
    layout(xaxis = list(title = 'bill_length_mm'))
```

# Interactive box plots (Video: [YouTube](https://youtu.be/VHxxzeJyCRY), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=d7ef9599-80fa-4f45-be52-af460142cc44))

We create interactive box plots using two approaches.

In the code below, we:

1. Use **ggplot2** to create a box plot of `bill_length_mm` for each `species`. We associate `species` with the x-variable and `bill_length_mm` with the y-variable. We assign this plot the name `ggbox`.
1. Use `plotly::ggplotly` to make the graphic interactive.

The interactive graphic indicates the 5-number summary (min, Q1, median, Q3, max) of `bill_length_mm` for each `species`. It also indicates the value of any outlier. 

```{r, warning = FALSE}
ggbox <- 
  ggplot(penguins) +
  geom_boxplot(aes(x = species, y = bill_length_mm))
ggplotly(ggbox)
```

We can create a similar set of parallel box plots using **plotly**. Using the `plot_ly` function, we:

- Associating `species` with the `x` attribute
- Associate `bill_length_mm` with the `y` attribute\
- Specify `type = "box"`. 

The interactive graphic indicates the `species` and 5-number summary (min, Q1, median, Q3, max) of `bill_length_mm` for each box plot. 

```{r, warning = FALSE}
plot_ly(penguins,
        x = ~species,
        y = ~bill_length_mm,
        type = "box")
```

# Interactive violin plots (Video: [YouTube](https://youtu.be/AKGuWX1qUvQ), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=740663e8-e7b6-4b2d-9567-af460142d481))

We use two approaches to create interactive violin plots.

In the code below, we:

1. Use **ggplot2** to create a violin plot of `bill_length_mm` for each `species`. We assign this plot the name `ggvio`.
1. Use `plotly::ggplotly` to make the graphic interactive.

The interactive graphic indicates the `species`, `bill_length_mm`, and the associated density for that value of `bill_length_mm` when we hover over a violin curve.

```{r, warning = FALSE}
ggvio <- 
  ggplot(penguins) +
  geom_violin(aes(x = species, y = bill_length_mm))
ggplotly(ggvio)
```

To create a similar plot, using the `plot_ly` function we:

- Associate `species` with the `x` attribute
- Associate `bill_length_mm` with the `y` attributes
- Specify `type = "violin"`.

The interactive graphic indicates the `species`, `bill_length_mm`, and the associated density for that value of `bill_length_mm` when we hover over a violin curve, as well as the 5-number summary of the associated box plot.

```{r, warning = FALSE}
plot_ly(penguins,
        x = ~species,
        y = ~bill_length_mm,
        type = "violin")
```

# Interactive scatter plots (Videos: [YouTube](https://youtu.be/umxR8dviHvg), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=cf34d7b8-ec05-47e0-914e-af4601895931))

We will create an interactive scatter plot of `bill_length_mm` versus `body_mass_g` for the `penguins` data that uses different colors and shapes to distinguish the different `species`. 

We will investigate two simple approaches for doing this.

In the code below, we:

1. Use **ggplot2** to create the grouped scatter plot. We assign this plot the name `ggscatter`.
1. Use `plotly::ggplotly` to make the graphic interactive.

The interactive graphic indicates the `species` (twice), `body_mass_g`, and `bill_length_mm` when we hover over a density curve.

```{r, warning = FALSE}
ggscatter <-
  ggplot(penguins) +
  geom_point(aes(x = body_mass_g,
                 y = bill_length_mm,
                 color = species,
                 shape = species))
ggplotly(ggscatter)
```

Notice that `species` is indicated twice when we hover over a point. We can correct this behavior by using the `tooltip` argument to specify the attributes (`x`, `y`, `color`, etc.) we want to display when our mouse hovers over a point.

```{r, warning = FALSE}
# restrict attributes displayed from hover
ggplotly(ggscatter,
         tooltip = c("shape", "x", "y"))
```

We can create a similar interactive scatter plot using `plot_ly`: We:

- Specify `type = "scatter"` and `mode = "marker"` to indicate that we want to plot points. 
- Associate `body_mass_g` with the `x` attribute and `bill_length_m`` with the `y` attribute for the actual points.
- Associate `species` with the `color` and `symbol` attributes to change those aspects of the plot.

The resulting scatter plot indicates the `species`, `body_mass_g`, and `bill_length_mm` of each point.

```{r, warning = FALSE}
plot_ly(penguins,
        x = ~body_mass_g,
        y = ~bill_length_mm,
        color = ~species,
        symbol = ~species,
        mode = "markers",
        type = "scatter")
```

# Interactive scatter plots with smooths (Video: [YouTube](https://youtu.be/dYocJTb5_Mc), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=35af8dfe-6bef-4021-b366-af46018a1711))

We now attempt to add some linear regression smooths to an interactive scatter plot.

In the code below, we:

1. Use **ggplot2** to create a scatter plot of `bill_length_mm` versus `body_mass_g` that uses different colors and shapes to distinguish the different `species`.
2. Add a second layer to the plot the provides an `"lm"` smooth for the points of each `species`. We assign this plot the name `ggsmooth`.
3. Use `plotly::ggplotly` to make the graphic interactive.

The interactive graphic indicates the `species`, `body_mass_g`, and `bill_length_mm` of each point, points on the .

```{r, warning = FALSE}
ggsmooth <- 
  ggplot(penguins) +
  geom_point(aes(x = body_mass_g,
                 y = bill_length_mm,
                 color = species,
                 shape = species)) +
  geom_smooth(aes(x = body_mass_g,
                  y = bill_length_mm,
                  color = species),
              method = "lm")
ggplotly(ggsmooth)
```

Adding a smooth to a plot using **plotly** natively is a bit more difficult because you have to manually compute the smooth, extract the fitted line for each group, and then add the fitted lines as a layer to an existing scatter plot.

In the code below, we fit a separate lines model for the data, which essentially fits a separate liner regression model to the points of each `species`. We then add the `fitted` values from this model as a new variable to the `penguins` data frame.

```{r, warning = FALSE}
# fit separate lines/interaction model
lmod <- lm(bill_length_mm ~ body_mass_g + species,
           data = penguins, na.action = na.exclude)
# add fitted values for each group
penguins$fitted <- fitted(lmod)
```

Now that all relevant data is in the `penguins` data frame, we:

- Use the same syntax as before to create a scatter plot of `bill_length_mm` versus `body_mass_g` that uses different colors and shapes for the points of each species.
- Use the native R pipe operator to add a "trace" to the original plot using `add_lines`.
    - We supply the `penguins` data frame to `add_lines` (make sure to specify `data = ` since that is not the first argument of the `add_lines` function).
    - Associate `body_mass_g` with the `x` attribute
    - Associate`fitted` with the `y` attribute.
    - Change the color of the lines by associated `species` with the `color` attribute.
    - Specify `inherit = FALSE`, which means we are not inheriting any of the attribute specifications in the `plot_ly` function (which are otherwise passed by default).
  
The interactive scatter plot allow us to see the `species`, `bill_length_mm`, and `body_mass_g` for each point and the points on the smoother line associated with each `species`.

```{r, warning = FALSE}
plot_ly(penguins,
        x = ~body_mass_g,
        y = ~bill_length_mm,
        mode = 'markers',
        color = ~species,
        symbol = ~species,
        type = 'scatter') |>
  add_lines(data = penguins,
    x = ~body_mass_g,
    y = ~fitted,
    color = ~species,
    inherit = FALSE)
```

# Interactive maps (Video: [YouTube](https://youtu.be/qrW2y5sQFBg), [Panopto](https://ucdenver.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=86d6de23-fe99-4981-aa0a-af46018a7775))

Interactive maps can provide a lot of information. We will create an interactive map using **ggplot2**, **plotly**, and the **sf** package [@R-sf]. 

First, we use the `st_read` function from the `sf` package to read a shapefile related to North Carolina packages that is installed by default with the `sf` package. The imported shapefile is automatically converted to an `sf` data frame. The imported object has many variables, but we point out three:

- `NAME`: the name of each North Carolina county
- `BIR74`: the number of recorded births in each county in 1974.
- `geometry`: the `MULTIPOLYGON` associated with each North Carolina county.

```{r, warning = FALSE}
# import sf object from shapefile in sf package
nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"),
                  quiet = TRUE)
# display first 3 rows of nc for certain variables
head(nc[c("NAME", "BIR74", "geometry")], n = 3)
```
In the code below, we:

1. Use **ggplot2** to create a choropleth map of `BIR74` for each county using `geom_sf`.
    a. We specify `fill = BIR74` so that the fill color of each county is based on the `BIR74` variable.
    b. We also associate the `NAME` variable with the `label` aesthetic so that the name of each county is displayed when we hover over a county.
    c. Use `scale_fill_viridis_c` to change the color palette used for the fill color.
    d. We assign this plot the name `ggsf`.
1. Use `plotly::ggplotly` to make the graphic interactive.

The interactive graphic indicates the number of births in each county and the county name when we hover over a county.

```{r, warning = FALSE}
# plot sf object using ggplot2
ggsf <-
  ggplot(nc) +
  geom_sf(aes(fill = BIR74, label = NAME)) +
  scale_fill_viridis_c()
# make map interactive
ggplotly(ggsf)
```

Is there a way to provide information from multiple variable simulatneously when we hover over a county? Yes! But we have to be creative. We:

- Use the `paste0` function to create a new variable, `info`, that combines multiple variables into a single character string for each county. The `\n` indicates to start a new line. We add a new line before each variable name.
- Add the `info` variable as a variable to the `nc` data frame.

```{r, warning = FALSE}
# combine multiple variables into a character string 
# (one per county)
info <- paste0(
  "\nname: ", nc$NAME,
  "\narea: ", nc$AREA,
  "\nbirths in 1974: ", nc$BIR74,
  "\nSIDS cases in 1974: ", nc$SID74)
# print first 2 values of info
info[1:2]
# add info the nc
nc$info <- info
```

Now, we use `info` as the `label` aesthetic in `geom_sf` and
specify `tooltip = "label"` so that only the `label` variable is displayed when we hover over a county.

```{r, warning = FALSE}
# create mape that fills based on BIR74 but the tooltip
# based on info
ggsf <-
  ggplot(nc) +
  geom_sf(aes(fill = BIR74, label = info)) +
  scale_fill_viridis_c()
# show only label tooltip
ggplotly(ggsf, tooltip = "label")
```

We can create a similar plot using `plot_ly`. We:

- Specify `type = "scatter"` and `mode = "lines"`.
- Associate the `info` variable in `nc` with the `split` attribute to draw the separate traces for each county. We could have used `NAME`, but then only the `NAME` of each county would be displayed when we hover. This way, we get additional information.
- Associate the `BIR74` variable in `nc` with the `color` attribute to fill each county with a color from a gradient.
- Specify `showlegend = FALSE` so that only the color scale is displayed and no legend related to `info`. **This is a critical step**.
- Specify `alpha = 1` so that the colors aren't muted.
- Specify `hoverinfo = "text"` so the only the `split` information is displayed
- Pipe this graphic into the `colorbar` function and change the title to "BIR74" (otherwise it gets displayed twice).

```{r, warning = FALSE}
plot_ly(nc,
        color = ~BIR74,
        split = ~info,
        showlegend = FALSE,
        alpha = 1,
        type = "scatter",
        mode = "lines",
        hoverinfo = "text")  |>
  colorbar(title = "BIR74")
```

# References
